\section{September 26, 2023}

\subsection{Total Variation Distance}

\begin{definition}
\deflabelname{Total Variation Distance}

Let $\mu$ and $\nu$ be two distributions on $\mathcal{X}$. The \ac{total variation distance}, $d_{TV}(\mu, \nu)$ is given by 
\[d_{TV}(\mu, \nu) = \sup_{A\subseteq \mathcal{X}}\vert \mu(A) - \nu(A)\vert.\]
\end{definition}

We can use $\mu$ and $\nu$ in the definition of $d_{TV}$ interchangeably with random variables distributed as $\mu$ and $\nu$ respectively. 

\begin{example}
\exlabel

If $X$ and $Y$ are Bernoulli random variables with parameters $p,q$ respectively, then $d_{TV}(X,Y) = \vert p-q\vert$. 
\end{example}

In this example, $\mathcal{X}=\{0,1\}$. For all possible $A\subseteq X$, the difference in their probability is at most $\vert p-q\vert$. 

\begin{theorem}
\proplabel

Total variation distance is a distance metric, along with some other properties:
\begin{itemize}
    \item $d_{TV}(\mu, \nu) = 0$
    \item $d_{TV}(\mu, \nu) = d_{TV}(\nu, \mu)$
    \item Triangle inequality: $d_{TV}(\mu, \nu)\leq d_{TV}(\mu, \eta) + d_{TV}(\eta, \nu)$
    \item 
    \[d_{TV}(\mu, \nu) = \frac{1}{2}\sum_{x\in \mathcal{X}}\vert \mu(x) - \nu(x)\vert = \sum_{x:\mu(x)>\nu(x)}\mu(x)-\nu(x).\]
    \item $X_n\cd X$ if and only if $d_{TV}(X_n, X)\rightarrow 0$. 
\end{itemize}
\end{theorem}

\begin{proof}
Proof of third bullet point: think of $\mu, \nu$ like a bar graph. Shade all area above $\mu$ and below $\nu$ red, and shade all area above $\nu$ and below $\mu$ blue. The maximal difference $\mu(A)-\nu(A)$ is achieved by collecting all the blue area, which is the expression on the right side. Since $\mu$ and $\nu$ are distributions, the red and blue areas are equal; since the middle expression is the sum of both areas, it is also equal to the right hand side. 
\end{proof}

The goal of defining a total variation distance is to eventually try to compute 
\[d_{TV}(\mu P^i, \pi).\]
As $i$ increases, the total variation distance to the stationary distribution should decrease. To help us understand this more concretely, we define a notion of \ac{coupling}. 

\begin{definition}
\deflabelname{Coupling}

A \ac{coupling} of two distributions $\mu$ and $\nu$ on probability space $\mathcal{X}$ is a joint distribution $\eta$ on $\mathcal{X}\times \mathcal{X}$ whose marginals are $\mu$ and $\nu$ respectively. 
\end{definition}

A coupling of random variables $X$ and $Y$ is a random variable $(\tilde{X}, \tilde{Y})$ for which $X\sim \tilde{X}$ and $Y\sim \tilde{Y}$. 

\begin{example}
\exlabel

Let $X$ and $Y$ be $\Bern(p)$ random variables. 
\end{example}

Then, the independent coupling is given by
\[\PP[(\tilde{X},\tilde{Y})=(x,y)] = 
\begin{cases}
(1-p)^2 & (x,y)=(0,0) \\
p(1-p) & (x,y)\in \{(0,1),(1,0)\} \\
p^2 & (x,y) = (1,1).
\end{cases}\]

Another coupling is to take $\tilde{X}=\tilde{Y}$:
\[\PP[(\tilde{X},\tilde{Y})=(x,y)] = 
\begin{cases}
1-p & (x,y)=(0,0) \\
p & (x,y)=(1,1) \\
0 & (x,y)\in \{(0,1),(1,0)\}.
\end{cases}\]

\begin{example}
\exlabel

Consider the finite Gambler's ruin MC with $n$ states. Let $X_0=x$, $Y_0=y$, with $x\leq y$. Show that for some $i$, $\PP[X_i=n|X_0=x]\leq \PP[Y_i=n|Y_0=y]$. 
\end{example}

We will use a coupling $(\tilde{X_i}, \tilde{Y_i})$ with $\tilde{X_0}=x$ and $\tilde{Y_0}=y$ that move left/right in parallel. This is a valid coupling, since the marginal distribution of each variable is the same as their individual distributions. Now, 

\[\PP[X_i=n] = \PP[\tilde{X_i}=n] = \PP[\tilde{X_i}=n, \tilde{Y_i}=n]\leq \PP[\tilde{Y_i}=n] = \PP[Y_i=n].\]

\begin{theorem}
\proplabel

$d_{TV}(\mu, \nu) \leq \inf\{\PP[X\neq Y] : (X,Y)\text{ is a coupling of }\mu, \nu\}.$
\end{theorem}

\begin{proof}
\[\mu(A) - \nu(A) = \PP[X\in A] - \PP[Y\in A] \leq \PP[X\in A, Y\neq A]\leq \PP[X\neq Y].\]
\end{proof}

This is always an inequality; there always exists coupling $(X,Y)$ such that $\PP[X\neq Y] = d_{TV}(\mu, \nu)$. We will not prove this, but one such example is to take $p\leq q$, $U\sim \Unif[0,1]$, $X=\mathbbm{1}_{u\leq p}$, $Y=\mathbbm{1}_{u\leq q}$. 

\begin{theorem}
\thmlabelname{Convergence Theorem using Coupling}

$P$ irreducible, aperiodic, $\vert \mathcal{X}\vert < \infty$. Then, we know there exists unique $\pi$ stationary, and 
\[X_i \cd \pi\]
for any starting distribution $X_0\sim \mu$. In other words, 
\[\lim_{i\rightarrow \infty}\PP[X_i=x|X_0\sim \mu] = \pi(x)\forall x.\]
\end{theorem}

\begin{proof}
	This is equivalent to showing that $d_{TV}(\mu P^i, \pi)\rightarrow 0$ as $i\rightarrow \infty$. Construct coupling $(X_i,Y_i)$ with $X_0\sim \mu$ and $Y_i\sim \pi$. Then, consider independently $X_i'$ and $Y_i'$ starting from $\mu, \pi$ respectively, and let $T$ be the first time that $X_i'=Y_i'$. Let $X_i=X_i'$ and $Y_i=Y_i'$ for all $i\leq T$ and $X_i=Y_i=X_i'$ for all $i > T$. This pairing has the correct marginal distributions, so it is a coupling. 

Now, we have 
\[d_{TV}(\mu P^i, \pi)\leq \PP[X_i\neq Y_i] = \PP[T > i].\]

Note that $(X_i,Y_i)$ is an MC on $\mathcal{X}\times \mathcal{X}$ with $(x,x)$ its only closed class. Therefore, $\PP[T>i]$ is the probability that we have not entered this closed class by time $i$, which approaches $0$ as $i\rightarrow \infty$, hence done. 
\end{proof}
